AttCRISPR : a spacetime interpretable model for prediction of sgRNA on-target activity

### 2020-11-13 Paper/main.tex
11-13 paper

\abstract{\textbf{Motivation:} More and more Cas9 variants with higher specificities are developed to avoid the off-target effect, which bring a significant volume of experimental data. 
Conventional machine learning performs poorly on these datasets, while the models based on deep learning are often lack of interpretability, 
which makes researchers have to trade-off accuracy and interpretability. 
We believe it is necessary to develope a model which can not only match deep learning based model in performance, 
but also with strong interpretability that can be comparable to conventional machine learning method.\\
\textbf{Results:} To overcome these problems, we design and implement AttCRISPR, an intrinsic interpretable method based on deep learning to predict the on-target activity. 
In this paper, we find that based on the preprocessing of the input base sequence, the deep learning models can be divided into two categories, 
encoded-based methods and embedded-based methods. 
Based on the categorization above, we assume encoded-based methods and embedded-based methods are heterogeneous, and use a full connection layer to stack them, which achieve a better performance. 
Our model is trained and tested on the DeepHF dataset for performance evaluation. 
AttCRISPR can compete with state-of-the-art method in these datasets, and achieves an average spearman value of 0.872, 0.867, 0.867 (corresponding to WT-SpCas9, eSpCas9(1.1), SpCas9-HF1) with the same evaluation strategy. 
In addition, benefits from two attention modules, spatial and temporal attention modules, AttCRISPR has strong interpretability. 
Through them, we can understand the decisions made by AttCRISPR at both global and local levels without other post hoc explanations techniques. 
With the trained model, we reveal the preference for each position-dependent nucleotide on the sgRNA sequence in each dataset at global level. 
And at local level, we prove that interpretability of AttCRISPR can be used to guide researcher to design sgRNA with higher activity, in order to reduce the experimental cost. 
\\
\textbf{Availability:} The example code are available at \href{https://github.com/South-Walker/AttCRISPR}{https://github.com/South-Walker/AttCRISPR}\\
\textbf{Contact:} \href{zrjiang@cs.ecnu.edu.cn}{zrjiang@cs.ecnu.edu.cn}\\
\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}
online.}

@@

In practice, there have been a number of application and toolkits applied in this task. In the earlier studies, the methods \emph{in silico} are categorized into three types: 
(1) alignment-based, (2) hypothesis-driven and (3) learning-based \citep{chuai2018deepcrispr}. 
Recently, we noticed that the last type of method seems to be getting more attention because of huger and huger datasets \citep{liu2019computational}.

@@

 In addition, the interpretability of existing models are all at global level, and result is a general pattern in the datasets. 
For example, which position has a great impact on the final decision of the model and which position-dependent nucleotide has a positive impact on the activity. 
Lack of analysis at local level. 
For example, which structure causes the low activity of a certain nucleotide sequence and how to improve its activity with a few modification. 
In light of the above, we believe it is necessary to develope a model which can not only match deep learning based model in performance, 
but also with strong interpretability that can be comparable to conventional machine learning method.

\begin{figure}[!tpb]%figure2
    \centerline{\includegraphics[width=86mm]{NEWcategory.png}}
    \caption{Two categories of the deep learning models used in sgRNA related task. 
    (a) Model work in spatial domain. 
    In spatial domain, the base sequence is encoded into a binary matrix (or a binary image). 
    Since convolution has great advantages in extracting spatial features, the convolutional neural network is an excellent tool in spatial domain. 
    (b) Model work in temporal domain. 
    In temporal domain, the base sequence (represented by the binary matrix) is embedded into a sequance of high-dimensional vector, 
    in which the RNN perform better.
    In addition, we note that the last layers of neural network are usually full connection structure (not necessarily), 
    which greatly increases the difficulty of understanding the decisions of model.
    }\label{fig:category}
\end{figure}

@@

\begin{enumerate}
    \item Methods in spatial domain. 
    Some previous studies have used model based CNN to predict sgRNA on-target activity or off-target effect \citep{chuai2018deepcrispr,kim2018deep,lin2018off-target}. 
    They process sgRNA base sequence inputs with the help of one-hot encoding idea. 
    In other words, they regard it as two-dimensional image data, and use convolution layer to extracte potential features in spatial domain, 
    It is worth noting that \citeauthor{zhang2020c-rnncrispr:} adds bidirectional gated recurrent unit (BGRU, in short), a RNN variant, after pooling layer of classic CNN network \citep{zhang2020c-rnncrispr:}. 
    Our explanation is that BGRU assists CNN to extract spatial features in one dimension, under this belief it belong to this category. 
    \item Methods in temporal domain. 
    Although RNN-based network have been shown effective to improve the performance of the model with temporal sequential input, especially in Natural Language Processing(NLP) and sequential recommendation, 
    RNN be not used for gRNA activity prediction, until recently \citep{Liu2019,wang2019optimized,liu2020deep}. 
    They consider the nucleotides(can also dimer or polymer) in the sgRNA sequence as word, and the sgRNA sequence itself as a sentence (from 5' to 3'), then a trainable matrix (could be either supervised or unsupervised) is used to project the word to the dense real-valued space. 
    This technology is called embedding, which generates the base embedding. RNN further encoding the base embedding into a sequence of hidden state vector. 
    Specially, \citeauthor{Liu2019} use RNN and CNN in parallel to extract features in base embedding \citep{Liu2019}. 
    However, base embedding is not spatially interpretable (different from one-hot encode), and they have no way to further explore the correlation between CNN and RNN output. 
    Almost all of the RNN based models used in sgRNA on-target activity or off-target effect flatten the hidden state vector into a one-dimensional vector as the input of the fully connected layer. 
    It is a pity that the temporal sequential dependency of hidden state vector are rarely noticed. 
    In summarize, RNN has limited representation power in capturing spatial feature. 
    Furthermore, the hidden state vector representation is usually hard to understand.
\end{enumerate}

Attention mechanism has demonstrated its power in NLP, Statistical Learning, Speech and Computer Vision. 
 It makes model tends to focus selectively on parts of the input, which is help in performing the task effectively. 
 Previous observation have shown that Cas9 preferentially binds sgRNAs containing purines but not pyrimidines \citep{wang2014genetic} and multiple thymine in the spacer impairing sgRNA activity \citep{wu2014genome-wide}, 
 that is to say some specific nucleotides and base position need more attention compared to others. 
 The above is the premise of introducing attention mechanism. Strictly speaking, we are not the first to bring attention mechanisms into this field. 
 The most similar approach to ours is the work based on transformer by \citeauthor{Liu2019}. 
 They use transformer, a components based on attention mechanism, instead of RNN to improve the ability of temporal feature extraction, 
 hence, enhance the performance of their model \citep{vaswani2017attention,Liu2019}. 
 In our work,the interpretability benefit from attention mechanism is more focused. 
 Our main contributions are as follows:\vspace*{1pt}
\begin{itemize}
    \item Present a novel deep-learning model, which can extract potential feature representation of sgRNA sequence in both spatial and temporal domain parallelly. 
    Finally, the ensemble learning method is used to combine the two to achieve better performance than current state-of-the-art models. 
    \item Introduce attention mechanism into our model. 
    As a result, it does not need post hoc explanations techniques based on input perturbation to explain itself. 
    It is intrinsic interpretable in both temporal and spatial domains.
    In the spatial domain it's at global level, while at local level in the temporal domain. 
    Thus, it is transformed from a black box to an intrinsically interpretable model with the performance of deep learning based model. 
    \item Throught ablation analysis and testing a series of possible network structures, 
    we find there are multiple components and strategy can improve the performance of AttCRISPR, 
    which could outperform current state-of-the-art tools on DeepHF dataset.\vspace*{1pt}
\end{itemize}

@@

For encoding process, we use the complementary base to represent the original base in sgRNA. 
Further, we use one-hot encode strategy, that is to say, we encoding each base in sgRNA into a four-dimensional vector 
(encode A,T,G,C into [1,0,0,0], [0,1,0,0], [0,0,1,0], [0,0,0,1], respectively), called one-hot vector. 
Then a sgRNA can be considered as a matrix $X_{oh}\in\mathbb{R}^{l\times4} $, named one-hot matrix (a little sparse, since an one-hot vector is zero in all but one dimension). 
We believe it is meaningful to regard $X_{oh}$ as a binary image, 
therefore, it is used as an input of convolutional neural network, which performs well in the image field.
Meanwhile, as mentioned above, one-hot matrix is a little sparse.

@@

Due to each element of $X_{oh}$ is interpretable (representing whether there is a corresponding nucleotide type at the corresponding location), 
we call $X_{oh}$ the spatial input, and the CNN work on $X_{oh}$ is method in spatial domain. 
In other hand, different from $X_{oh}$, $X_e$ can only be explained in the first dimension (representing the embedding vector of corresponding nucleotide type), 
and embedding vector is difficult for humans to understand. 
That's why we call $X_e$ the temporal input, and the RNN (or its variant) work on $X_e$ belongs to the method in temporal domain.

@@

Based on the categorization above, we assume that the method in spatial domain and in temporal domain are heterogeneous, 
which can satisfy the diversity premise of ensemble learning. 
Based on assumption above and ensemble learning, we follow the stacking strategy to develop AttCRISPR 
which can extract potential feature representation of sgRNA sequence in both spatial and temporal domain parallelly. 
Further, we apply attention mechanisms in both spatial and temporal domain to enhance the interpretability of AttCRISPR.

@@

In order to introduce the neural network architecture of AttCRISPR, Let's define first-order preference and second-order preference for convenience. 
Taking a simple linear regression model as an example, for input $X$ in Equation~(\ref{eq:01}), $y$ in Equation~(\ref{eq:01}) is as following
 \begin{equation}
y = AX\label{eq:linear}
\end{equation}
where $A \in\mathbb{R}^d$, the total differential of $y$ in Equation~(\ref{eq:linear}) is as following
 \begin{equation}
dy = \sum^{d}_iA_idX_i\label{eq:02}
\end{equation}
where $A_i$ and $X_i$ denotes the $i$-th dimension of the vector $X$ and $A$, 
$A_i$ indicates how dramatically the function changes as $X_i$ changes in a neighborhood of $X$, 
in other word, the importance of $X_i$. 
That's why we'll call $A$ first-order preference in our paper. 
Specifically, we use a vector $A_i$ to build the first-order original preference at position $i$ within sgRNA sequence, 
and $X_i$ is an embeddedness of the $i$-th feature, then $A$ and $X$ are two matrix. 
Further, the final result can be weighted by a trainable non-negative weight vector $W\in\mathbb{R}^l$, as follow
\begin{equation}
y=W \cdot AX^{T}\label{eq:superlinear}
\end{equation}
then we define $\tilde{A}$ as the first-order combine preference matrix (or just first-order preference), which means $\tilde{A}$ can be expressed linearly by $A$ as follow
\begin{equation}
\tilde{A}=BA\label{eq:04}
\end{equation}
where the weight matrix $B\in\mathbb{R}^{l\times l}$ is learned through attention mechanism, 
which we'll call the second-order preference matrix in our paper due to its calculation is based on first-order preference, 
it can explain how a particular pattern containing two nucleotides affects the base sequence. 
Then the predicted value can be expressed as
\begin{equation}
y=W \cdot \tilde{A} {X_e}^{T}\label{eq:05}
\end{equation}

@@

\begin{figure}[!tpb]%figure2
    \centerline{\includegraphics[width=86mm]{CNNv2.png}}
    \caption{The architecture of spatial domain method in AttCRISPR. 
    The input of the method is encoded sgRNA sequence $X_{oh}$, a $21\times 4$ one-hot matrix. 
    Then refine it through a spatial attention module, which could tell us the importance of a specific matrix element (or just say, pixel).
    A simple CNN followed is applied to extract potential feature representation of sgRNA sequence.
    In the last step, we flatten the output of CNN network into a one dimensional vector and 
    use a multilayer perceptron with sigmoid activation function to achieve the spatial output $y_s$.}\label{fig:CNN}
\end{figure}
As Figure~\ref{fig:CNN} demonstrated, the method in spatial domain relies on the convolutional neural network. 
As previously mentioned, sgRNA sequence has been encoded into a $21\times 4$ one-hot matrix $X_{oh}$, and we regard $X_{oh}$ as a binary image. 
Then, convolution kernels with different size are used to extract potential spatial features just like what other do in computer vision. 
According to the foregoing, the spatial attention Module proposed by \citeauthor{woo2018cbam:} can be applied in our method, 
which has been used to improve the performance of CNN in vision tasks.

@@

\subsubsection{Model ensemble following stacking strategy}

Some indirect sgRNA features, which can't be obtained directly by deep learning, 
including position accessibilities of secondary structure, stem–loop of secondary structure, melting temperature, 
and GC content are strongly associated with sgRNA activity \citep{Wang2016ProteinSS,wang2019optimized}. 
It's worth noting that in the work of \citeauthor{wang2019optimized}, hand-crafted biological features didn't be normalized. 
Since the wide range of data distribution, we believe it makes sense to normalize it, and we preprocess they based on the standard score (Z-Score).

@@

\subsection{Current prediction method}

@@

Two different sets of experiments are carried out in our work. 
The first one is designed for ablation analysis of AttCRISPR. 
We compare the performance of end2end method (without any hand-crafted biological features) in both spatial and temporal domain. 
Furthermore, we test the ensemble method based on the same strategy to prove that the ensemble of method in both spatial and temporal domain can significantly improve the performance.

The second experiment is designed to compare the performance of AttCRISPR and other current prediction method. 
In order to make the comparison apples to apples, we reduce the dimensionality of the same hand-crafted biological features as DeepHF's, 
which has been shown to enhance the predictability of a deep-learning model greatly, with a multilayer perceptron. 
Then follow the Equation~(\ref{eq:20}) to achieve the final prediction value. 
AttCRISPR (with the hand-crafted biological features) performs better on all three datasets than DeepHF.

\begin{table}[!tpb]
    \processtable{The methods compared in our work and brief description of these methods\label{Tab:baseline}} 
    {\begin{tabular}{@{}lllp{3.5cm}l@{}}\toprule
        Method & Neural & End2end & Description\\\midrule
        CNN$^*$ & Yes & Yes & Naive convolutional neural network\\
        RNN$^*$ & Yes & Yes & Bidirectional long short-term memory neural network\\
        XGBoost$^*$ & No & Yes & Extreme Gradient Boosting regression tree\\
        MLP$^*$ & Yes & Yes & Multilayer perceptron\\
        DeepHF$^*$ & Yes & No & Bidirectional long short-term memory neural network (with hand-crafted biological features)\\
        CRISPRpred(SEQ)$^{\#}$ & No & No & A conventional machine learning pipeline\\
        SpAC & Yes & Yes & Spatial AttCRISPR\\
        TAC & Yes & Yes & Temporal AttCRISPR\\
        EnAC & Yes & Yes & Ensemble AttCRISPR (without hand-crafted biological features)\\
        StAC & Yes & No & Standard AttCRISPR\\
        \botrule
    \end{tabular}}\footnotesize\setlength{\parindent}{2em}{\emph{Note}: The method with superscript of $^*$ and $^{\#}$ is reported by \citeauthor{wang2019optimized} and \citeauthor{MuhammadRafid2020} respectively. 
Specially, CRISPRpred(SEQ) take another set of hand-crafted sequence-based feature to improve performance.}
\end{table}

@@

We designed experiments to address the following questions:
\begin{itemize}
    \item 
    In the absence of hand-crafted biological features, whether the stacking of method in spatial domain and temporal domain can get better performance then use these method alone? 
    $\rightarrow$ Section~\ref{section:stacking}
    \item 
    How does AttCRISPR perform compared to current state-of-the-art methods, covering both conventional machine learning and deep-learning models? 
    $\rightarrow$ Section~\ref{section:comparison}
    \item 
    How can researcher understand the decisions made by AttCRISPR locally and globally, based on attention mechanisms? 
    $\rightarrow$ Section~\ref{section:interpretability}
    \vspace*{1pt}
\end{itemize}

@@

In order to make the comparison apples to apples, 
we follow Equation~(\ref{eq:20}) to modify the ensemble method and design the control experiment using the same strategy to validate the conclusion that integrating with hand-crafted biological features can improve the predictive performance of methods. 
What's more, we compare the standard AttCRISPR and DeepHF (Table~\ref{Tab:withbiofeature}).

As shown in Table~\ref{Tab:withbiofeature}, in the absence of hand-crafted biological features, AttCRISPR has significant advantages over DeepHF in predictability. 
Further, integration with the hand-crafted biological features can also improve the performance of AttCRISPR, and achieve Spearman correlation coefficients of 0.872, 0.867 and 0.867 for WT-SpCas9, eSpCas9(1.1) and SpCas9-HF1, respectively. Meanwhile, DeepHF achieve 0.867, 0.862 and 0.860 respectively. 
After integrating with biological features, the performance gap between AttCRISPR and DeepHF is shortened, while AttCRISPR still has better performance. In addition, we also compare the standard deviation of data obtained in ten tests, which are also shown in Table~\ref{Tab:withbiofeature}. 
It reveal that AttCRISPR is more stable than DeepHF.

@@

At global level, a important question we expect AttCRISPR to answer is which nucleotide it prefers at each position on the sequence. 
In fact, \citeauthor{wang2019optimized} has already answered this question in detail with the DeepSHAP method.
Leaving aside nucleotides, just consider the importance of each position, the method of \citeauthor{Liu2019} have gone some way too. 
While our method is not based on the post hoc explanations techniques and input perturbations, the only work we need to do is to get the first-order preference $A$ generated by the attention module.
Specifically, we use the first-order preference $A_s$ generated by the spatial attention module instead of  $\tilde{A}$ generated by the temporal attention module. 
The latter is in a higher dimensional dense space which make it difficult to understand. 
In practice, we input every sgRNA into the spatial AttCRISPR, in order to obtain the $A_s$ from the spatial attention module and take its mean value. 
Then we rescale it through Z-score to obtain a standardization value and the final result is shown in Figure~\ref{fig:spatialattention}.

As shown in Figure~\ref{fig:spatialattention}, we captured the preference for each position-dependent nucleotide on the sgRNA sequence. 
The result revealed that A and G typically have a positive contribution to the activity of sgRNA, while T typically have a negative contribution. 
It is agree with the previous conlusion that when Cas9 is binding sgRNA, it prefer the one containing purines to pyrimidines \citep{wang2014genetic}. 
In addition, global interpretability also pointed out that distinct from other nucleotides, G is strongly favored at position 20. 
This is consistent with the conclusions of several other reports \citep{Wong2015,Doench2014}.

>>>>

\abstract{\textbf{Motivation:} More and more Cas9 variants with higher specificities are developed to avoid the off-target effect, which bring a significant volume of experimental data. 
Conventional machine learning performs poorly on these datasets, while the methods based on deep learning are often lack of interpretability, 
which makes researchers have to trade-off accuracy and interpretability. 
It is necessary to develope a method which can not only match deep learning based methods in performance, 
but also with strong interpretability that can be comparable to conventional machine learning methods.\\
\textbf{Results:} To overcome these problems, we design and implement AttCRISPR, an intrinsic interpretable method based on deep learning to predict the on-target activity. 
The advantage of AttCRISPR lies in using the ensemble learning strategy to stack available encoded-based methods and embedded-based methods with strong interpretability. 
Comparison with the state-of-the-art methods (WT-SpCas9, eSpCas9(1.1), SpCas9-HF1), AttCRISPR can achieve an average spearman value of 0.872, 0.867, 0.867, respectively on several public datasets, which is superior to these methods. 
Furthermore, AttCRISPR has strong interpretability benefits from two attention modules, spatial and temporal attention modules. 
Through them, we can understand the decisions made by AttCRISPR at both global and local levels without other post hoc explanations techniques. 
With the trained model, we reveal the preference for each position-dependent nucleotide on the sgRNA (small guide RNA) sequence in each dataset at global level. 
And at local level, we prove that interpretability of AttCRISPR can be used to guide researcher to design sgRNA with higher activity. \\
\textbf{Availability:} The example code are available at \href{https://github.com/South-Walker/AttCRISPR}{https://github.com/South-Walker/AttCRISPR}\\
\textbf{Contact:} \href{zrjiang@cs.ecnu.edu.cn}{zrjiang@cs.ecnu.edu.cn}\\
\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}
online.}

@@

In practice, there have been a number of applications and toolkits applied in this task. In the earlier studies, the methods \emph{in silico} are categorized into three types: 
(1) alignment-based, (2) hypothesis-driven and (3) learning-based \citep{chuai2018deepcrispr}. 
Recently, we noticed that the last type of method seems to be getting more attention because of huger and huger datasets \citep{liu2019computational}.

@@

 In addition, the interpretability of existing models are all at global level, and result is a general pattern in the datasets. 
For example, which position has a great impact on the final decision of the model and which position-dependent nucleotide has a positive impact on the activity. 
Lack of analysis at local level. 
For example, which structure causes the low activity of a certain nucleotide sequence and how to improve its activity with a few modification. 
In light of the above, we believe it is critical to develop an effective model which can not only have good performance, but also with strong interpretability.

\begin{figure}[!tpb]%figure2
    \centerline{\includegraphics[width=86mm]{NEWcategory.png}}
    \caption{Two categories of the deep learning models used in sgRNA related task. 
    (a) Model work in spatial domain. 
    In spatial domain, the base sequence is encoded into a binary matrix (or a binary image). 
    Since convolution has great advantages in extracting spatial features, the CNN is an excellent tool in spatial domain. 
    (b) Model work in temporal domain. 
    In temporal domain, the base sequence (represented by the binary matrix) is embedded into a sequance of high-dimensional vector, 
    in which the RNN perform better.
    In addition, we note that the last layers of neural network are usually full connection structure (not necessarily), 
    which greatly increases the difficulty of understanding the decisions of model.
    }\label{fig:category}
\end{figure}

@@

\begin{enumerate}
    \item Methods in spatial domain. 
    Some previous studies have used the methods based CNN to predict sgRNA on-target activity or off-target effect \citep{chuai2018deepcrispr,kim2018deep,lin2018off-target}. 
    They process sgRNA base sequence inputs with the help of one-hot encoding idea. 
    In other words, they regard it as two-dimensional image data, and use convolution layer to extracte potential features in spatial domain, 
    It is worth noting that \citeauthor{zhang2020c-rnncrispr:} adds bidirectional gated recurrent unit (BGRU, in short), a RNN variant, after pooling layer of classic CNN network \citep{zhang2020c-rnncrispr:}. 
    Our explanation is that BGRU assists CNN to extract spatial features in one dimension, under this belief it belong to this category. 
    \item Methods in temporal domain. 
    Although RNN-based network have been shown effective to improve the performance of the model with temporal sequential input, especially in Natural Language Processing(NLP) and sequential recommendation, 
    RNN be not used for gRNA activity prediction, until recently \citep{Liu2019,wang2019optimized,liu2020deep}. 
    They consider the nucleotides(can also dimer or polymer) in the sgRNA sequence as word, and the sgRNA sequence itself as a sentence (from 5' to 3'), then a trainable matrix (could be either supervised or unsupervised) is used to project the word to the dense real-valued space. 
    This technology is called embedding, which generates the base embedding. RNN further encoding the base embedding into a sequence of hidden state vector. 
    Specially, \citeauthor{Liu2019} use RNN and CNN in parallel to extract features in base embedding \citep{Liu2019}. 
    However, base embedding is not spatially interpretable (different from one-hot encode), and they have no way to further explore the correlation between CNN and RNN output. 
    Almost all of the RNN based models used in sgRNA on-target activity or off-target effect flatten the hidden state vector into a one-dimensional vector as the input of the fully connected layer. 
    It is a pity that the temporal sequential dependency of hidden state vector are rarely noticed. 
    In summarize, RNN has limited representation power in capturing spatial feature. 
    Furthermore, the hidden state vector representation is usually hard to understand.
\end{enumerate}

Attention mechanism has demonstrated its power in NLP, Statistical Learning, Speech and Computer Vision. 
 It makes model tends to focus selectively on parts of the input, which is help in performing the task effectively. 
 Previous observation have shown that Cas9 preferentially binds sgRNAs containing purines but not pyrimidines \citep{wang2014genetic} and multiple thymine in the spacer impairing sgRNA activity \citep{wu2014genome-wide}, 
 that is to say some specific nucleotides and base position need more attention compared to others. 
 The above is the premise of introducing attention mechanism. Strictly speaking, we are not the first to bring attention mechanisms into this field. 
 The most similar approach to ours is the work based on transformer by \citeauthor{Liu2019}. 
 They use transformer, a components based on attention mechanism, instead of RNN to improve the ability of temporal feature extraction, 
 hence, enhance the performance of their model \citep{vaswani2017attention,Liu2019}. 
 In our work,the interpretability benefit from attention mechanism is more focused. 
 Our main contributions are as follows:\vspace*{1pt}
\begin{itemize}
    \item Present a novel deep-learning model, which can extract potential feature representation of sgRNA sequence in both spatial and temporal domain parallelly. 
    Finally, the ensemble learning method is used to combine the two to achieve better performance than current state-of-the-art models. 
    \item Introduce attention mechanism into our model. 
    As a result, it does not need post hoc explanations techniques based on input perturbation to explain itself. 
    It is intrinsic interpretable in both temporal and spatial domains.
    In the spatial domain it's at global level, while at local level in the temporal domain. 
    Thus, it is transformed from a black box to an intrinsically interpretable model with the performance of deep learning based model. 
    \item Throught ablation analysis and testing a series of possible network structures, 
    we find there are multiple components and strategies can improve the performance of AttCRISPR, 
    which could outperform current state-of-the-art tools on DeepHF dataset.\vspace*{1pt}
\end{itemize}

@@

For encoding process, we use the complementary base to represent the original base in sgRNA. 
Further, we use one-hot encode strategy, that is to say, we encoding each base in sgRNA into a four-dimensional vector 
(encode A,T,G,C into [1,0,0,0], [0,1,0,0], [0,0,1,0], [0,0,0,1], respectively), called one-hot vector. 
Then a sgRNA can be considered as a matrix $X_{oh}\in\mathbb{R}^{l\times4} $, named one-hot matrix (a little sparse, since an one-hot vector is zero in all but one dimension). 
We believe it is meaningful to regard $X_{oh}$ as a binary image, 
therefore, it is used as an input of CNN, which performs well in the image field.
Meanwhile, as mentioned above, one-hot matrix is a little sparse.

@@

Due to each element of $X_{oh}$ is interpretable (representing whether there is a corresponding nucleotide type at the corresponding location), 
we call $X_{oh}$ the spatial input, and the CNN work on $X_{oh}$ is the method in spatial domain. 
In other hand, different from $X_{oh}$, $X_e$ can only be explained in the first dimension (representing the embedding vector of corresponding nucleotide type), 
and embedding vector is difficult for humans to understand. 
That's why we call $X_e$ the temporal input, and the RNN (or its variant) work on $X_e$ belongs to the method in temporal domain.

@@

Based on the categorization above, we assume that the method in spatial domain and in temporal domain are heterogeneous, 
which can satisfy the diversity premise of ensemble learning. 
Based on the assumption above and ensemble learning, we follow the stacking strategy to develop AttCRISPR 
which can extract potential feature representation of sgRNA sequence in both spatial and temporal domain parallelly. 
Further, we apply attention mechanisms in both spatial and temporal domain to enhance the interpretability of AttCRISPR.

@@

In order to introduce the neural network architecture of AttCRISPR, Let's define first-order preference and second-order preference for convenience. 
Taking a simple linear regression model as an example, for input $X$ in Equation~(\ref{eq:01}), $y$ in Equation~(\ref{eq:01}) is as following
 \begin{equation}
y = AX\label{eq:linear}
\end{equation}
where $A \in\mathbb{R}^d$, the total differential of $y$ in Equation~(\ref{eq:linear}) is as following
 \begin{equation}
dy = \sum^{d}_iA_idX_i\label{eq:02}
\end{equation}
where $A_i$ and $X_i$ denotes the $i$-th dimension of the vector $X$ and $A$, 
$A_i$ indicates how dramatically the function changes as $X_i$ changes in a neighborhood of $X$, 
in other words, the importance of $X_i$. 
That's why we'll call $A$ first-order preference in our paper. 
Specifically, we use a vector $A_i$ to build the first-order original preference at position $i$ within sgRNA sequence, 
and $X_i$ is an embeddedness of the $i$-th feature, then $A$ and $X$ are two matrix. 
Further, the final result can be weighted by a trainable non-negative weight vector $W\in\mathbb{R}^l$, as follow
\begin{equation}
y=W \cdot AX^{T}\label{eq:superlinear}
\end{equation}
then we define $\tilde{A}$ as the first-order combine preference matrix (or just first-order preference), which means $\tilde{A}$ can be expressed linearly by $A$ as follow
\begin{equation}
\tilde{A}=BA\label{eq:04}
\end{equation}
where the weight matrix $B\in\mathbb{R}^{l\times l}$ is learned through attention mechanism, 
which we'll call the second-order preference matrix in our paper due to its calculation is based on first-order preference, 
it can explain how a particular pattern containing two nucleotides affects the base sequence. 
Then the predicted value can be expressed as
\begin{equation}
y=W \cdot \tilde{A} {X_e}^{T}\label{eq:05}
\end{equation}

@@

\begin{figure}[!tpb]%figure2
    \centerline{\includegraphics[width=86mm]{CNNv2.png}}
    \caption{The architecture of spatial domain method in AttCRISPR. 
    The input of the method is encoded sgRNA sequence $X_{oh}$, a $21\times 4$ one-hot matrix. 
    Then refine it through a spatial attention module, which could tell us the importance of a specific matrix element (or just say, pixel).
    A simple CNN followed is applied to extract potential feature representation of sgRNA sequence.
    In the last step, we flatten the output of CNN network into a one dimensional vector and 
    use a multilayer perceptron with sigmoid activation function to achieve the spatial output $y_s$.}\label{fig:CNN}
\end{figure}
As Figure~\ref{fig:CNN} demonstrated, the method in spatial domain relies on the CNN. 
As previously mentioned, sgRNA sequence has been encoded into a $21\times 4$ one-hot matrix $X_{oh}$, and we regard $X_{oh}$ as a binary image. 
Then, convolution kernels with different size are used to extract potential spatial features just like what other do in computer vision. 
According to the foregoing, the spatial attention module proposed by \citeauthor{woo2018cbam:} can be applied in our method, 
which has been used to improve the performance of CNN in vision tasks.

@@

\subsubsection{Ensemble model following stacking strategy}

Some indirect sgRNA features, which can't be obtained directly by deep learning, 
including position accessibilities of secondary structure, stem–loop of secondary structure, melting temperature, 
and GC content are strongly associated with sgRNA activity \citep{Wang2016ProteinSS,wang2019optimized}. 
It's worth noting that in the work of \citeauthor{wang2019optimized}, hand-crafted biological features didn't be normalized. 
Since the wide range of data distribution, we normalize it based on the standard score (Z-Score).

@@

\subsection{Current prediction methods}

@@

Two different experiments are carried out in our work. 
The first one is designed for ablation analysis of AttCRISPR. 
We compare the performance of end2end method (without any hand-crafted biological features) in both spatial and temporal domain. 
Furthermore, we test the ensemble method based on the same strategy to prove that the ensemble method in both spatial and temporal domain can significantly improve the performance.

The second experiment is designed to compare the performance of AttCRISPR with other current prediction methods. 
In order to make the comparison apples to apples, we reduce the dimensionality of the same hand-crafted biological features as DeepHF's, 
which has been shown to enhance the predictability of a deep-learning model greatly, with a multilayer perceptron. 
Then follow the Equation~(\ref{eq:20}) to achieve the final prediction value. 
AttCRISPR (with the hand-crafted biological features) performs better on all three datasets than DeepHF.

\begin{table}[!tpb]
    \processtable{The methods compared in our work and brief description of these methods\label{Tab:baseline}} 
    {\begin{tabular}{@{}lllp{3.5cm}l@{}}\toprule
        Method & Neural & End2end & Description\\\midrule
        CNN$^*$ & Yes & Yes & Naive CNN\\
        RNN$^*$ & Yes & Yes & Bidirectional long short-term memory neural network\\
        XGBoost$^*$ & No & Yes & Extreme Gradient Boosting regression tree\\
        MLP$^*$ & Yes & Yes & Multilayer perceptron\\
        DeepHF$^*$ & Yes & No & Bidirectional long short-term memory neural network (with hand-crafted biological features)\\
        CRISPRpred(SEQ)$^{\#}$ & No & No & A conventional machine learning pipeline\\
        SpAC & Yes & Yes & Spatial AttCRISPR\\
        TAC & Yes & Yes & Temporal AttCRISPR\\
        EnAC & Yes & Yes & Ensemble AttCRISPR (without hand-crafted biological features)\\
        StAC & Yes & No & Standard AttCRISPR\\
        \botrule
    \end{tabular}}\footnotesize\setlength{\parindent}{2em}{\emph{Note}: The method with superscript of $^*$ and $^{\#}$ is reported by \citeauthor{wang2019optimized} and \citeauthor{MuhammadRafid2020} respectively. 
Specially, CRISPRpred(SEQ) take another set of hand-crafted sequence-based feature to improve performance.}
\end{table}

@@

We design experiments to address the following questions:
\begin{itemize}
    \item 
    In the absence of hand-crafted biological features, whether the stacking of method in spatial domain and temporal domain can get better performance then use these method alone? 
    $\rightarrow$ Section~\ref{section:stacking}
    \item 
    How does AttCRISPR perform compared to current state-of-the-art methods, covering both conventional machine learning and deep-learning models? 
    $\rightarrow$ Section~\ref{section:comparison}
    \item 
    How can researcher understand the decisions made by AttCRISPR locally and globally, based on attention mechanisms? 
    $\rightarrow$ Section~\ref{section:interpretability}
    \vspace*{1pt}
\end{itemize}

@@

In order to validate the conclusion that integrating with hand-crafted biological features can improve the predictive performance of methods, 
we follow Equation~(\ref{eq:20}) to modify the ensemble method and design the control experiment using the same strategy. 
What's more, we compare the standard AttCRISPR and DeepHF (Table~\ref{Tab:withbiofeature}).

As shown in Table~\ref{Tab:withbiofeature}, in the absence of hand-crafted biological features, AttCRISPR has significant advantages over DeepHF in predictability. 
Further, integration with the hand-crafted biological features can also improve the performance of AttCRISPR, and achieve Spearman correlation coefficients of 0.872, 0.867 and 0.867 for WT-SpCas9, eSpCas9(1.1) and SpCas9-HF1, respectively. Meanwhile, DeepHF achieve 0.867, 0.862 and 0.860, respectively. 
After integrating with biological features, the performance gap between AttCRISPR and DeepHF is shortened, while AttCRISPR still has better performance. In addition, we also compare the standard deviation of data obtained in ten tests, which are also shown in Table~\ref{Tab:withbiofeature}. 
It reveal that AttCRISPR is more stable than DeepHF.

@@

At global level, an important question we expect AttCRISPR to answer is which nucleotide it prefers at each position on the sequence. 
In fact, \citeauthor{wang2019optimized} has already answered this question in detail with the DeepSHAP method.
Leaving aside nucleotides, just consider the importance of each position, the method of \citeauthor{Liu2019} have gone some way too. 
While our method is not based on the post hoc explanations techniques and input perturbations, the only work we need to do is to get the first-order preference $A$ generated by the attention module.
Specifically, we use the first-order preference $A_s$ generated by the spatial attention module instead of  $\tilde{A}$ generated by the temporal attention module. 
The latter is in a higher dimensional dense space which make it difficult to understand. 
In practice, we input every sgRNA into the spatial AttCRISPR, in order to obtain the $A_s$ from the spatial attention module and take its mean value. 
Then we rescale it through Z-score to obtain a standardization value and the final result is shown in Figure~\ref{fig:spatialattention}.

As shown in Figure~\ref{fig:spatialattention}, we captured the preference for each position-dependent nucleotide on the sgRNA sequence. 
The result revealed that A and G typically have a positive contribution to the activity of sgRNA, while T typically have a negative contribution. 
It is agree with previous conlusion that when Cas9 is binding sgRNA, it prefer the one containing purines to pyrimidines \citep{wang2014genetic}. 
In addition, global interpretability also pointed out that distinct from other nucleotides, G is strongly favored at position 20. 
This is consistent with the conclusions of several other reports \citep{Wong2015,Doench2014}.
